---
title: Which Package a Dataset Comes From
author: Samuel Bohman
date: "2017-10-03"
tags: 
  - r
  - r-packages 
slug: which-package-a-dataset-comes-from
output: blogdown::html_page
summary: "Have you ever wondered where a particular dataset is coming from?"
---

Have you ever wondered where a particular dataset is coming from? Yesterday, I read a blog post in which the author used a dataset called "gold", but he did not mention the origin of the dataset. To find out, I developed a simple function called `datasearch`. It takes one argument, `p`, which is a character string containing a regular expression `pattern` (used by `grepl`) and returns a `data.frame`: 

```{r, warning = FALSE}
datasearch <- function(p) {
  ds <- as.data.frame(data(package = .packages(all.available = TRUE))$results)
  ds[grepl(p, ds$Package) | grepl(p, ds$Item) | grepl(p, ds$Title), -2]
}

datasearch("gold")
```

```{r, warning = FALSE}
datasearch("1982")
```

At the moment, the function only searches the packages that are available on the local computer, not on [CRAN](https://cran.r-project.org/) or [GitHub](https://github.com/).

Also useful:

```{r}
# list of datasets in a particular package
data(package = "ggplot2")
```

