---
title: Jaccard similarity function for binary attributes
author: Samuel Bohman
date: '2017-09-14'
tags:
  - r
  - jaccard-similarity
  - jaccard-distance
slug: jaccard-similarity-function-for-binary-attributes
output:
  blogdown::html_page
summary: "Function for calculating the Jaccard similarity and distance coefficients for binary attributes."
---

## Introduction

The Jaccard similarity is a distance function which measures the similarity between two sets of data. In the simplest case where we have  binary attributes, meaning the attributes are either 0 or 1, true or false, etc., the Jaccard similarity is equal to the intersection of A and B divided by their union: 

$$
J_S(A, B) = \frac{|A \cap B|}{|A \cup B|} = \frac{|A \cap B|}{|A| + |B| - |A \cap B|}
$$

The complement to the Jaccard similarity is the Jaccard distance and it measures dissimilarity between two sets: 

$$
J_D(A, B) = 1 - J_S(A, B)
$$

### Example 

For example, Assume we have two vectors A and B. To calculate the Jaccard similarity we use the following formula: 

$$
J = \frac{M_{11}}{M_{01} + M_{10} + M_{11}}
$$

In other words, the Jaccard similarity coefficient measures the number of attributes where A and B are both 1, divided by the number of attributes where A and B are dissimilar, plus the number of attributes where they are both 1. 

{{% alert note %}}
Notice that the Jaccard similarity does not include the combination where they are both 0.  
{{% /alert %}}

## The Function `jaccard.R`

The function calcuates the Jaccard similarity and the Jaccard distance for binary attributes. It takes two arguments: `x` a dataframe or matrix object, and `m` the `MARGIN` argument used in the `apply` function. If your data is in long format set `m = 1` to apply `sum` over the rows. Conversely, If your data is in wide format set `m = 2` to apply `sum` over the columns. Logical vectors work too. 

```{r, echo = FALSE}
blogdown::shortcode("gist", "samuel-bohman", "e0df06467661da6023a4eab6ab06e791")
```

## Examples

Data in long format: 

```{r}
df1 <- data.frame(
  IDS = c(1, 1, 1, 1, 1, 1, 1, 1, 1, 0), 
  CESD = c(1, 1, 1, 0, 1, 1, 0, 0, 0, 0))
head(df1)
```

```{r}
source("jaccard.R")
jaccard(df1, 1)
```

Data in wide format: 

```{r}
df2 <- data.frame(Q1 = c(1L, 0L), Q2 = c(0L, 1L), Q3 = c(0L, 1L), Q4 = c(1L, 0L), Q5 = c(1L, 1L))
df2
```

```{r}
jaccard(df2, 2)
```

Logical vectors work too: 

```{r}
df3 <- data.frame(A = c(TRUE, TRUE, TRUE), B = c(TRUE, TRUE, FALSE))
df3
```

```{r}
jaccard(df3, 1)
```

Enjoy! `r emo::ji("smile")`
