---
title: Distance Metrics
author: Samuel Bohman
date: "2017-10-05"
tags: 
  - distance-metrics
  - euclidean
  - manhattan
  - minkowski
slug: distance-metrics
output: blogdown::html_page
summary: "A brief overview of distance metrics used in unsupervised learning."
---

There are many distance metrics; here we will take a look at three: the Euclidean distance, the Manhattan distance, and the Minkowski distance.  

### Euclidean distance

The Euclidean or direct "bird-flight" distance is by far the most commonly used distance metric. It can be thought of as the straight line distance between two points. It is defined as follows: 

$$
\left( \sum_{j=1}^P \left(x_{aj} - x_{bj}\right)^2 \right)^{1/2}
$$

#### Example

Assume we have two vectors: $X = [2, 2, 3, -1]$ and $Y = [1, 4, 4, 0]$. The Euclidean distance between $X$ and $Y$ can then be calculated as follows: 

$$L_2(x, y) = \sqrt{(2-1)^2 + (2-4)^2 + (3-4)^2 + (-1-0)^2}$$

$$=\sqrt{1^2 + (-2)^2 + (-1)^2 + (-1)^2}$$

$$=\sqrt{1 + 4 + 1 + 1}$$

$$=\sqrt{7} \approx 2.65$$

In R: 

```{r}
x <- c(2, 2, 3, -1)
y <- c(1, 4, 4, 0)
m <- matrix(data = c(x, y), byrow = T, nrow = 2)
dist(m, method = "euclidean")
```

### Manhattan Distance

The Manhattan or "city-block" distance (or "Pac-Man" distance `r emo::ji("smile")`) is another common metric used for datasets with binary predictors. 

```{r, echo = FALSE}
blogdown::shortcode("figure", 
                    src = "https://upload.wikimedia.org/wikipedia/commons/0/08/Manhattan_distance.svg", 
                    alt = "Manhattan distance")
```

In the figure, the blue, red, and yellow lines show various ways of getting from the lower left corner to the upper right corner using the grid layout. The green straight line shows the Euclidean distance in comparison. 

The Manhattan metric is defined as follows: 

$$
L_1(x, y) = \sum_{i = 1}^d | x_i - y_i |
$$

#### Example 

Assume we have two vectors: $X = [0, 5, 5, 6, 6]$ and $Y = [0, 0, 2, 2, 6]$ which corresponds to the yellow line in the figure. Thus, the Manhattan distance between $X$ and $Y$ can then be calculated as follows: 

$$L_1(x, y) = |0-0| + |5-0| + |5-2| + |6-2| + |6-6| = 12$$
In R, we can do the following: 

```{r}
# The yellow line
x <- c(0, 5, 5, 6, 6)
y <- c(0, 0, 2, 2, 6)
z <- matrix(c(x, y), byrow = T, nrow = 2)
dist(z, method = "manhattan")
```

### Minkowski Distance

The Minkowski distance can be considered a generalization of both the Euclidean distance and the Manhattan distance. It is defined as follows: 

$$
L_p(x, y) = \left( \sum_{i = 1}^d | x_i - y_i |^p \right)^{1/p}
$$
Minkowski distance is typically used with $p$ being 1 or 2, which correspond to $L_1$ or the Manhattan distance and $L_2$ or the Euclidean distance, respectively. 
